---
title: "Sports Analytics Project"
author: "Ian Keller"
date: "2023-10-19"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading in Each CSV
```{r}
advanced_stats = read.csv('Player Season Totals - 2022-23 Advanced Stats.csv')

basic_stats = read.csv('Player Season Totals - 2022-23 Basic Stats.csv')

player_bio = read.csv('Player Season Totals - 2022-23 Player Bios.csv')
```

Merging Data Sets Based on Name

```{r}
#Combining the player bio and basic stats data sets
combined_data <- merge(player_bio, basic_stats, by = "Player")

#Combining the advanced stats with the previously merged data
combined_data2 <- merge(advanced_stats, combined_data, by = "Player")
```

Adding in the salary dataset
```{r}
player_salary = read.csv("2022-2023 Player Salaries.csv")
#Indexing only the player name and salary columns
player_and_salary <- player_salary[, c("Name", "AAV")]

#Renaming the "Name" Column to "Player" for a proper merge
library("dplyr")
player_and_salary <- player_and_salary %>%
  rename(Player = Name)

#Renaming the player columns to match case sensitivity
combined_data2$Player <- tolower(combined_data2$Player)
player_and_salary$Player <- tolower(player_and_salary$Player)

#Finally merging all the data into the final dataset
final_data <- merge(combined_data2, player_and_salary, by = "Player")
```


Next, I will have to clean the data. Remove irrelevant and duplicate columns as we move closer towards analysis.
-For starters, I want to remove some count variables like Corsi For and Corsi Against and leave Corsi %
```{r}
library(dplyr)
cleaned_data <- final_data %>% select(-CF,-CA,-FF,-FA,-SF,-SA,-GF,-GA,-xGF,-xGA,-SCF,-SCA,-HDCF,-HDCA,-HDGF,-HDGA, -MDCF,-MDCA,-MDGF,-MDGA,-LDCF,-LDCA,-LDGF,-LDGA,-Off..Zone.Starts,-Neu..Zone.Starts,-Def..Zone.Starts,-On.The.Fly.Starts,-Off..Zone.Faceoffs,-Neu..Zone.Faceoffs,-Def..Zone.Faceoffs,-X.x,-Team.x,-Position.x,-Birth.Country,-Birth.City,-Date.of.Birth,-Draft.Team,-Round.Pick,-X.y,-Team.y,-Position.y,-GP.y,-X,-Faceoffs.Won,-Faceoffs.Lost,-Hits.Taken,-iCF,-iFF,-iSCF,-iHDCF,-SH.,-ixG,-Total.Penalties,-Minor,-Major,-Misconduct,-FF,-TOI.y,-Team, -On.Ice.SH.,-On.Ice.SV.,-HDCF.,-Nationality,-Birth.State.Province)
```
I now want to convert some count statistics to rate variables. Additionally, I want to create a variable called "Years_since_draft" as I believe the longer they have been in the league can correlate to higher salaries. I expect this variable to be normally distributed as well, meaning that the middle of the career is the highest with two tails at a young and old age.
```{r}
library(dplyr)

#Converting Draft.Year collumn to a numeric value
cleaned_data$Draft.Year <- as.numeric(cleaned_data$Draft.Year)

cleaned_data2 <- cleaned_data %>%
  mutate(
    GPG = Goals / GP.x,
    TOIpg = TOI.x / GP.x,
    AsPG = Total.Assists / GP.x,
    Primary_AsPG = First.Assists / GP.x,
    Secondary_AsPG = Second.Assists / GP.x,
    Years_since_draft = abs(Draft.Year - 2022),
    PPG = Total.Points / GP.x)
```
Testing if there are any duplicate names
```{r}
library(dplyr)

non_unique_players <- cleaned_data2 %>%
  group_by(Player) %>%
  filter(n() > 1) %>%
  arrange(Player)

print(unique(non_unique_players$Player))
```
There were 16 duplicate sebastian aho rows, but only two players in the league with that name. This will filter it out to show the two proper rows.
```{r}
library(dplyr)

non_unique_players <- non_unique_players[c(3, 14), ]

```
Merging the two datasets
```{r}
other_players_data <- cleaned_data2 %>%
  filter(Player != "sebastian aho")

# Combine the two datasets
all_data_filtered <- bind_rows(non_unique_players, other_players_data)
```

Removing all NA values for salary and renaming AAV to Salary
```{r}
all_data_filtered <- na.omit(all_data_filtered, cols = "AAV")
#Renaming AAV
all_data_filtered <- all_data_filtered %>%
  rename(Salary = AAV)
```

Furthermore, I need to transform Salary from a string to an integer
```{r}
library(readr)

all_data_filtered$Salary <- parse_number(all_data_filtered$Salary)
```
Multiplying Salary by 1,000,000
```{r}
all_data_filtered$Salary <- all_data_filtered$Salary * 1000000
```

Removing salaries of 0
```{r}
library(dplyr)

all_data_filtered <- all_data_filtered %>%
  filter(Salary != 0 & GP.x >= 41)
```

Looking at some descriptive statistics. First let's inspect the dependent variable of salary (AAV).

```{r}
library(ggplot2)
library(scales)

ggplot(all_data_filtered, aes(x = Salary)) +
  geom_histogram(bins = 10, fill = "darkgrey", color = "white") +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma, breaks = scales::pretty_breaks(n = 12)) +
  labs(title = "Histogram of Salary", x = "Salary (in dollars)", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```
Performing a log transformation
```{r}
all_data_filtered$Log_salary <- log(all_data_filtered$Salary)
```
Histogram of Log Salaries
```{r}
library(ggplot2)
library(scales)

ggplot(all_data_filtered, aes(x = Log_salary)) +
  geom_histogram(bins = 10, fill = "darkgrey", color = "white") +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma, breaks = scales::pretty_breaks(n = 12)) +
  labs(title = "Histogram of Salary Transformed", x = "Salary (Log Transformed)", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


5 number summary
```{r}
# Calculate the 5-number summary for the Salary variable
five_num_summary <- summary(all_data_filtered$Salary)
print(five_num_summary)
```
Creating a binary column for Offense vs Defense
```{r}
all_data_filtered$Offense_defense <- ifelse(all_data_filtered$Position == "D",1,0)
```
Plotting Points Per Game against Salary
```{r}
model <- lm(Salary / 1e6 ~ PPG, data = all_data_filtered)
plot(all_data_filtered$PPG, all_data_filtered$Salary / 1e6, pch = 19, xlab = "Points Per Game (5v5)", ylab = "Salary (Millions of dollars)", main = "Points Per Game vs Salary")
abline(model, col = "black")
```
Creating a subset of data for a correlation plot
```{r}
library(dplyr)
subset <- all_data_filtered %>%
  ungroup() %>%
  select(Salary, TOIpg, Overall.Draft.Position, PPG, Offense_defense, Years_since_draft)
```
Creating the correlation plot
```{r}
library(corrplot)
subset2 <- data.frame(lapply(subset, as.numeric))
#Transforming Draft Position to a numeric value before creating the model
all_data_filtered$Overall.Draft.Position <- as.numeric(all_data_filtered$Overall.Draft.Position)
# Calculate the correlation matrix
correlation_matrix = cor(subset2)
corrplot(correlation_matrix, method = "color", addCoef.col = "black")
```

Creating a multiple regression model for our hypothesized significant variables
```{r}
hypothesized_best_model <- lm(Log_salary ~ TOIpg + Overall.Draft.Position + PPG + Offense_defense + Years_since_draft, data = all_data_filtered)
summary(hypothesized_best_model)
```
Using all subsets regression to find the best possible model using all of our predictors
```{r}
#First I will need to create a subset of the data frame
full_predictors <- all_data_filtered[c(
  "Log_salary", "CF.", "FF.", "SF.", "xGF.", "SCF.", "PDO", "Draft.Round",
  "Overall.Draft.Position", "IPP", "Shots", "Rush.Attempts", "Rebounds.Created", "PIM",
  "Penalties.Drawn", "Giveaways", "Takeaways", "Hits", "Shots.Blocked", "GPG", "TOIpg",
  "AsPG", "Years_since_draft", "Offense_defense", "GF.", "HDGF.", "MDGF.", "LDGF.", "PPG"
)]

#Making the entire df numeric
library(dplyr)

full_predictors <- full_predictors %>%
  mutate_all(as.numeric)

```
Running the all subsets regression based on adjr2
```{r}
require(MASS)

min_model = lm(Log_salary ~ 1, data = full_predictors)
max_model = lm(Log_salary ~ CF. + FF. + SF. + xGF. + SCF. + PDO + Draft.Round + Overall.Draft.Position + IPP + Shots + Rush.Attempts + Rebounds.Created + PIM + Penalties.Drawn + Giveaways + Takeaways + Hits + Shots.Blocked + GPG + TOIpg + AsPG + Years_since_draft + Offense_defense + GF. + HDGF. + MDGF. + LDGF. + PPG, data = full_predictors)

best_model = step(min_model, scope = formula(max_model), direction = "both", k = log(538))

```
Summary Statistics and ANOVA table of the best model
```{r}
best_BIC_model = lm(Log_salary ~ AsPG + Years_since_draft + Shots + TOIpg + Offense_defense + 
    Overall.Draft.Position + Draft.Round, data = full_predictors)

summary(best_BIC_model)
anova(best_BIC_model)
```
Plotting the hypothesized and best BIC model residuals
```{r}
plot(full_predictors$Log_salary, residuals(hypothesized_best_model), pch =19, main = "Residual Plot of Hypothesized Best Model", xlab ="Fitted Values", ylab = "Residuals")
abline(h=0,col="red")
```
```{r}
plot(full_predictors$Log_salary, residuals(best_BIC_model), pch =19, main = "Residual Plot of BIC Best Model", xlab ="Fitted Values", ylab = "Residuals")
abline(h=0,col="red")
```
Running the full model
```{r}
full_model = lm(Log_salary ~ PPG + CF. + FF. + SF. + xGF. + SCF. + PDO + Draft.Round + Overall.Draft.Position + IPP + Shots + Rush.Attempts + Rebounds.Created + PIM + Penalties.Drawn + Giveaways + Takeaways + Hits + Shots.Blocked + GPG + TOIpg + Years_since_draft + Offense_defense + GF. + HDGF. + MDGF. + LDGF., data = full_predictors)
summary(full_model)
```


